---
title: "Customizing Combined Plots"
author: "Ronald (Ryy) G. Thomas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customizing Combined Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  dpi = 96,
  fig.retina = 1,
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 4.5
)
```

```{r load, message=FALSE}
library(zzlongplot)
library(ggplot2)
library(patchwork)
```

## Overview

When `lplot()` is called with `plot_type = "both"`, it returns a
patchwork object that places the observed values and
change-from-baseline plots side by side. This object can be
modified after creation using two patchwork operators:

- `&` applies a modification to **all** panels
- `+` applies a modification to the **last** panel only

This vignette demonstrates common post-creation customizations.

## Example data

```{r data}
set.seed(42)
n <- 20

trial <- data.frame(
  subject_id = rep(1:n, each = 4),
  visit = rep(0:3, times = n),
  arm = rep(c("Drug", "Placebo"), each = 4 * n / 2)
)
trial$score <- 50 +
  ifelse(trial$arm == "Drug", -3, -0.5) * trial$visit +
  rnorm(nrow(trial), sd = 4)
```

## Base combined plot

```{r base}
p <- lplot(trial, score ~ visit | arm, baseline_value = 0,
           plot_type = "both")
p
```

## Prefer an argument where one exists

Post-composition is the right tool for anything patchwork-specific,
and for one-off adjustments. It is the wrong tool for settings
`lplot()` already accepts, for two reasons: an added scale replaces
whatever scale the plot already carried, and an added **complete**
theme (`theme_bw()`, `theme_minimal()`, a journal theme) resets every
theme setting `lplot()` made, including the plot margin and legend
position that the sample-size table depends on. A partial
`theme(...)` call, which changes only the elements it names, is safe.

Several arguments exist precisely so that common customizations do not
have to go through `+`:

| Instead of appending | Pass |
|:---------------------|:-----|
| `scale_x_continuous(breaks = ...)` | `x_breaks` |
| `labs(colour = ..., fill = ..., linetype = ..., shape = ...)` | `legend_title` |
| `theme_bw(base_size = 9)` | `base_size` |
| `facet_wrap(..., labeller = ...)` | `facet_type`, `facet_labeller` |
| `geom_point(size = ...)`, `geom_line(linewidth = ...)` | `point_size`, `line_width` |

`legend_title` is worth singling out. The grouping column is renamed
internally, so retitling the legend by hand means naming all four
aesthetics: under `bw_print` the linetype and shape are mapped to the
group as well as the color, and retitling only the color splits one
key into two.

```{r args-not-composition}
lplot(trial, score ~ visit | arm, baseline_value = 0,
      plot_type = "obs",
      x_breaks = 0:3,
      legend_title = "Treatment arm",
      point_size = 2.5,
      line_width = 0.8,
      base_size = 11,
      title = "Set through arguments, not post-composition")
```

## Modifying all panels with `&`

The `&` operator passes a ggplot2 element to every panel in the
composition. This is useful for global theme changes, font
adjustments, or color overrides.

### Change the theme

```{r theme}
p & theme_classic()
```

### Adjust font size

```{r font}
p & theme(text = element_text(size = 14))
```

### Override group colors

```{r colors}
p & scale_color_manual(values = c("Drug" = "#0072B2",
                                  "Placebo" = "#D55E00"))
```

### Remove gridlines

```{r grid}
p & theme(panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank())
```

## Modifying a single panel with `+`

The `+` operator targets only the last panel (the change plot in
a `"both"` layout). This allows selective modifications when the
two panels need different treatment.

### Add a reference line to the change plot

```{r refline}
p + geom_hline(yintercept = 0, linetype = "dashed",
               color = "grey50")
```

## Adding overall annotations

The `plot_annotation()` function from patchwork adds titles,
subtitles, captions, and tag labels that span the entire
composition, sitting above or below the individual panel labels.

```{r annotation}
p + plot_annotation(
  title = "Efficacy Results: Drug vs Placebo",
  subtitle = "Study 001, ITT Population",
  caption = "Error bars represent standard error of the mean.",
  tag_levels = "A"
)
```

Panel tags (`tag_levels = "A"`) label each panel as (A), (B),
etc., which is a common requirement for journal submissions.

### Styling the annotation

Annotation text inherits from the plot theme but can be
overridden:

```{r annotation-style}
p + plot_annotation(
  title = "Efficacy Results",
  tag_levels = "A",
  theme = theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.tag = element_text(face = "bold")
  )
)
```

## Combining multiple modifications

The operators can be chained. Use `&` for global changes first,
then `+` for composition-level annotations.

```{r combined}
(p &
  theme_minimal() &
  theme(legend.position = "bottom",
        text = element_text(size = 12))) +
  plot_annotation(
    title = "Phase III Trial: Primary Endpoint",
    tag_levels = "A"
  )
```

## Controlling layout

The layout itself can be adjusted after creation using
`plot_layout()`:

```{r layout}
p + plot_layout(widths = c(2, 1))
```

This produces a wider observed panel and a narrower change panel,
which can be useful when the change plot carries less visual
complexity.

## Summary

| Operator | Scope | Example |
|:---------|:------|:--------|
| `&` | All panels | `p & theme_bw()` |
| `+` | Last panel (or annotation) | `p + geom_hline(...)` |
| `+ plot_annotation()` | Entire composition | titles, tags, captions |
| `+ plot_layout()` | Entire composition | widths, heights, guides |

Because `lplot()` returns a standard patchwork object, any
technique documented in the
[patchwork package](https://patchwork.data-imaginist.com/)
applies directly.

One caution when combining the two approaches: `&` with a complete
theme replaces the theme `lplot()` applied, so a figure using
`sample_size_opts = list(position = "table")` loses the margin
reserved for its count rows. Either set the type size with
`base_size` rather than re-theming, or use `region = "panel"`, which
places the rows inside the panel and needs no reserved margin.
