---
title: "Customizing Combined Plots"
author: "Ronald (Ryy) G. Thomas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customizing Combined Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 4.5
)
```

```{r load, message=FALSE}
library(zzlongplot)
library(ggplot2)
library(patchwork)
```

## Overview

When `lplot()` is called with `plot_type = "both"`, it returns a
patchwork object that places the observed values and
change-from-baseline plots side by side. This object can be
modified after creation using two patchwork operators:

- `&` applies a modification to **all** panels
- `+` applies a modification to the **last** panel only

This vignette demonstrates common post-creation customizations.

## Example data

```{r data}
set.seed(42)
n <- 20

trial <- data.frame(
  subject_id = rep(1:n, each = 4),
  visit = rep(0:3, times = n),
  arm = rep(c("Drug", "Placebo"), each = 4 * n / 2)
)
trial$score <- 50 +
  ifelse(trial$arm == "Drug", -3, -0.5) * trial$visit +
  rnorm(nrow(trial), sd = 4)
```

## Base combined plot

```{r base}
p <- lplot(trial, score ~ visit | arm, baseline_value = 0,
           plot_type = "both")
p
```

## Modifying all panels with `&`

The `&` operator passes a ggplot2 element to every panel in the
composition. This is useful for global theme changes, font
adjustments, or color overrides.

### Change the theme

```{r theme}
p & theme_classic()
```

### Adjust font size

```{r font}
p & theme(text = element_text(size = 14))
```

### Override group colors

```{r colors}
p & scale_color_manual(values = c("Drug" = "#0072B2",
                                  "Placebo" = "#D55E00"))
```

### Remove gridlines

```{r grid}
p & theme(panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank())
```

## Modifying a single panel with `+`

The `+` operator targets only the last panel (the change plot in
a `"both"` layout). This allows selective modifications when the
two panels need different treatment.

### Add a reference line to the change plot

```{r refline}
p + geom_hline(yintercept = 0, linetype = "dashed",
               color = "grey50")
```

## Adding overall annotations

The `plot_annotation()` function from patchwork adds titles,
subtitles, captions, and tag labels that span the entire
composition, sitting above or below the individual panel labels.

```{r annotation}
p + plot_annotation(
  title = "Efficacy Results: Drug vs Placebo",
  subtitle = "Study 001, ITT Population",
  caption = "Error bars represent standard error of the mean.",
  tag_levels = "A"
)
```

Panel tags (`tag_levels = "A"`) label each panel as (A), (B),
etc., which is a common requirement for journal submissions.

### Styling the annotation

Annotation text inherits from the plot theme but can be
overridden:

```{r annotation-style}
p + plot_annotation(
  title = "Efficacy Results",
  tag_levels = "A",
  theme = theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.tag = element_text(face = "bold")
  )
)
```

## Combining multiple modifications

The operators can be chained. Use `&` for global changes first,
then `+` for composition-level annotations.

```{r combined}
(p &
  theme_minimal() &
  theme(legend.position = "bottom",
        text = element_text(size = 12))) +
  plot_annotation(
    title = "Phase III Trial: Primary Endpoint",
    tag_levels = "A"
  )
```

## Controlling layout

The layout itself can be adjusted after creation using
`plot_layout()`:

```{r layout}
p + plot_layout(widths = c(2, 1))
```

This produces a wider observed panel and a narrower change panel,
which can be useful when the change plot carries less visual
complexity.

## Summary

| Operator | Scope | Example |
|:---------|:------|:--------|
| `&` | All panels | `p & theme_bw()` |
| `+` | Last panel (or annotation) | `p + geom_hline(...)` |
| `+ plot_annotation()` | Entire composition | titles, tags, captions |
| `+ plot_layout()` | Entire composition | widths, heights, guides |

Because `lplot()` returns a standard patchwork object, any
technique documented in the
[patchwork package](https://patchwork.data-imaginist.com/)
applies directly.
