---
title: "The Formula Interface"
author: "Ronald (Ryy) G. Thomas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Formula Interface}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r load, message=FALSE}
library(zzlongplot)
library(ggplot2)
```

## Overview

The `lplot()` function uses a formula interface to specify the
relationship between outcome, time, and grouping variables. This
design follows the convention established by base R modeling
functions such as `lm()` and `aov()`, making the syntax familiar
to most R users.

The general form is:

```
outcome ~ time | group
```

where each component maps to a plot element:

- **outcome** (left of `~`): the response variable, plotted on the
  y-axis
- **time** (right of `~`, before `|`): the independent variable,
  plotted on the x-axis
- **group** (after `|`): an optional grouping variable, mapped to
  color and linetype

The formula is parsed internally by `parse_formula()`, which
extracts each component as a character string for downstream use
in summary statistics and plot construction.

## Example data

The examples below use a simulated longitudinal dataset with two
treatment arms measured at five visits.

```{r data}
set.seed(123)
n <- 20

trial <- data.frame(
  subject_id = rep(1:n, each = 5),
  visit = rep(0:4, times = n),
  arm = rep(c("Drug", "Placebo"), each = 5 * n / 2)
)
trial$score <- 50 +
  ifelse(trial$arm == "Drug", -2.5, -0.5) * trial$visit +
  rnorm(nrow(trial), sd = 4)
```

## Simple formula: `y ~ x`

The minimal formula specifies only an outcome and a time variable.
This produces a single line with error bars summarising all
subjects together, with no group differentiation.

```{r simple}
lplot(trial, score ~ visit, baseline_value = 0,
      plot_type = "obs")
```

## Adding a group: `y ~ x | group`

The pipe operator `|` introduces a grouping variable. Each level
of the group receives its own line and color.

```{r grouped}
lplot(trial, score ~ visit | arm, baseline_value = 0,
      plot_type = "obs")
```

## Multiple grouping variables: `y ~ x | g1 + g2`

When the data contain more than one factor of interest, combine
them with `+`. The groups are concatenated into a single
interaction factor for plotting.

```{r multi-group}
trial$sex <- rep(c("Male", "Female"), length.out = nrow(trial))

lplot(trial, score ~ visit | arm + sex, baseline_value = 0,
      plot_type = "obs")
```

## Adding facets: `y ~ x | group` with `facet_form`

Faceting is specified through a separate `facet_form` argument
rather than through the main formula. This keeps the primary
formula concise while still supporting panel layouts.

```{r facet}
lplot(trial, score ~ visit | arm, baseline_value = 0,
      facet_form = ~ sex,
      plot_type = "obs")
```

For two-dimensional faceting, use a two-sided formula:

```{r facet-grid, eval=FALSE}
lplot(trial, score ~ visit | arm, baseline_value = 0,
      facet_form = sex ~ site,
      plot_type = "obs")
```

## Baseline auto-detection

When `baseline_value` is omitted (or set to `NULL`), `lplot()`
attempts to identify the baseline visit automatically. For numeric
visit variables, the minimum value is used. For character or
factor visit codes, the function searches for common labels
including `'bl'`, `'BL'`, `'baseline'`, `'screening'`, `'scr'`,
`'day 0'`, `'week 0'`, `'pre'`, and `'visit 1'`.

A message is printed to the console reporting which value was
selected.

```{r auto-detect}
lplot(trial, score ~ visit | arm, plot_type = "obs")
```

If the visit variable uses a recognised character code, the
detection works the same way:

```{r auto-detect-char}
trial_char <- trial
trial_char$visit <- factor(
  trial_char$visit,
  levels = 0:4,
  labels = c("bl", "wk4", "wk8", "wk12", "wk16")
)

lplot(trial_char, score ~ visit | arm, plot_type = "obs")
```

If no common baseline label is found, or if multiple candidates
match (e.g., a dataset containing both `'bl'` and `'baseline'`),
the function raises an informative error asking the user to set
`baseline_value` explicitly.

## Change from baseline

The `plot_type` argument controls whether observed values, change
scores, or both are displayed. The formula itself does not change;
the same specification drives all three plot types.

```{r change}
lplot(trial, score ~ visit | arm, baseline_value = 0,
      plot_type = "change")
```

```{r both, fig.width=10, fig.height=4.5}
lplot(trial, score ~ visit | arm, baseline_value = 0,
      plot_type = "both")
```

## Formula parsing details

The `parse_formula()` function is exported and can be called
directly to inspect how a formula is decomposed:
```{r parse}
parse_formula(score ~ visit | arm)
```

For formulas with faceting specified in the main formula (an
alternative syntax), a second `~` introduces facet variables:

```{r parse-facet}
parse_formula(score ~ visit | arm ~ site)
parse_formula(score ~ visit | arm ~ site + region)
```

## Summary

The table below lists the supported formula patterns and their
corresponding plot mappings.

| Formula | y-axis | x-axis | Color/group | Facet |
|:--------|:-------|:-------|:------------|:------|
| `y ~ x` | y | x | none | none |
| `y ~ x \| g` | y | x | g | none |
| `y ~ x \| g1 + g2` | y | x | g1:g2 | none |
| `y ~ x \| g ~ f` | y | x | g | f |
| `y ~ x \| g ~ f1 + f2` | y | x | g | f1, f2 |

The `facet_form` argument provides an equivalent and often clearer
way to specify facets separately from the main formula:

| Main formula | facet_form | Result |
|:-------------|:-----------|:-------|
| `y ~ x \| g` | `~ f` | columns by f |
| `y ~ x \| g` | `f1 ~ f2` | grid: rows by f1, columns by f2 |
