---
title: "MMRM Analysis with zzlongplot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MMRM Analysis with zzlongplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
```

```{r setup, include = FALSE}
library(zzlongplot)
library(ggplot2)
```

# Introduction

Mixed models for repeated measures (MMRM) are the standard primary
analysis method for continuous longitudinal endpoints in clinical
trials. Regulatory agencies (FDA, EMA) routinely expect MMRM as the
primary or sensitivity analysis in confirmatory studies. The
`zzlongplot` package integrates MMRM model fitting (via the `mmrm`
package) with `emmeans` pairwise contrasts, embedding the results
directly into longitudinal plots as significance annotations.

This vignette demonstrates the `test_method = "mmrm"` option and its
interaction with covariance structure selection, p-value adjustment,
and visualization.

# Simulated clinical trial data

We simulate a two-arm parallel-group trial with four post-baseline
visits. The active arm shows a treatment effect emerging at visit 2.

```{r data}
set.seed(42)
n_subj <- 80

df <- data.frame(
  subject_id = rep(1:n_subj, each = 5),
  visit = rep(c(0, 4, 8, 12, 16), times = n_subj),
  group = rep(
    c("Placebo", "Active"), each = 5,
    length.out = n_subj * 5
  )
)

df$measure <- rnorm(nrow(df), mean = 50, sd = 10)
df$measure[df$group == "Active" & df$visit >= 8] <-
  df$measure[df$group == "Active" & df$visit >= 8] - 7
```

# Basic MMRM plot

Setting `test_method = "mmrm"` fits a mixed model for repeated
measures with an interaction between treatment group and visit, then
extracts pairwise contrasts at each timepoint via `emmeans`.
Significance stars appear above points where the adjusted p-value
crosses the standard thresholds.

```{r basic-mmrm, fig.cap = "MMRM with unstructured covariance (default)."}
lplot(df, measure ~ visit | group,
  baseline_value = 0,
  cluster_var = "subject_id",
  statistical_annotations = TRUE,
  test_method = "mmrm",
  title = "MMRM: Unstructured Covariance (default)")
```

By default (`cov_struct = "auto"`), the package selects unstructured
covariance for studies with 10 or fewer timepoints and compound
symmetry otherwise. If the chosen structure fails to converge, it
falls back through progressively simpler structures (heterogeneous
Toeplitz, heterogeneous AR(1), compound symmetry) before giving up.

# Covariance structure selection

The covariance structure models the within-subject correlation
pattern across visits. The choice can affect both model fit and
inference. Common options:

- **`"us"`** (unstructured): estimates a separate variance for each
  visit and a separate correlation for each visit pair. The most
  flexible option and the regulatory default. Requires the most
  parameters: t(t+1)/2 for t timepoints.

- **`"cs"`** (compound symmetry): assumes equal variance across
  visits and equal correlation between all visit pairs. Only 2
  parameters. Appropriate when all timepoints are exchangeable.

- **`"ar1"`** (first-order autoregressive): assumes equal variance
  and exponentially decaying correlation with increasing time lag.
  Only 2 parameters. Appropriate for equally spaced visits where
  nearby measurements are more correlated.

- **`"ar1h"`** (heterogeneous AR(1)): AR(1) correlation structure
  but allows each visit to have its own variance. Good when
  variability changes over time (common in dose-titration studies).

- **`"toep"`** / **`"toeph"`** (Toeplitz / heterogeneous Toeplitz):
  correlation depends only on the lag between visits, not which
  specific visits are compared. A middle ground between AR(1) and
  unstructured.

- **`"sp_exp"`** (spatial exponential): similar to AR(1) but
  accommodates unequally spaced visits naturally.

```{r cov-ar1, fig.cap = "AR(1) covariance structure."}
lplot(df, measure ~ visit | group,
  baseline_value = 0,
  cluster_var = "subject_id",
  statistical_annotations = TRUE,
  test_method = "mmrm",
  cov_struct = "ar1",
  title = "MMRM: AR(1) Covariance")
```

```{r cov-cs, fig.cap = "Compound symmetry covariance structure."}
lplot(df, measure ~ visit | group,
  baseline_value = 0,
  cluster_var = "subject_id",
  statistical_annotations = TRUE,
  test_method = "mmrm",
  cov_struct = "cs",
  title = "MMRM: Compound Symmetry")
```

# P-value adjustment

The `p_adjust_method` parameter controls multiple comparison
correction. For MMRM, this is passed to `emmeans::contrast()` as the
`adjust` argument.

```{r adjust-none, fig.cap = "No multiplicity adjustment."}
lplot(df, measure ~ visit | group,
  baseline_value = 0,
  cluster_var = "subject_id",
  statistical_annotations = TRUE,
  test_method = "mmrm",
  p_adjust_method = "none",
  title = "MMRM: No p-value Adjustment")
```

```{r adjust-bonf, fig.cap = "Bonferroni adjustment."}
lplot(df, measure ~ visit | group,
  baseline_value = 0,
  cluster_var = "subject_id",
  statistical_annotations = TRUE,
  test_method = "mmrm",
  p_adjust_method = "bonferroni",
  title = "MMRM: Bonferroni Adjustment")
```

# Three-arm trial

MMRM with `emmeans` handles multi-arm trials naturally. For three or
more groups, the omnibus test (via `emmeans::joint_tests()`) is
reported at each timepoint, and pairwise contrasts are drawn as
bracket annotations below the omnibus p-value. Only significant
pairs (after any multiplicity adjustment) are displayed, keeping
the figure uncluttered when many comparisons are non-significant.

```{r three-arm-data}
set.seed(123)
n3 <- 90
df3 <- data.frame(
  subject_id = rep(1:n3, each = 5),
  visit = rep(c(0, 4, 8, 12, 16), times = n3),
  group = rep(
    c("Placebo", "Low Dose", "High Dose"), each = 5,
    length.out = n3 * 5
  )
)
df3$measure <- rnorm(nrow(df3), mean = 50, sd = 10)
df3$measure[df3$group == "Low Dose" & df3$visit >= 8] <-
  df3$measure[df3$group == "Low Dose" & df3$visit >= 8] - 4
df3$measure[df3$group == "High Dose" & df3$visit >= 8] <-
  df3$measure[df3$group == "High Dose" & df3$visit >= 8] - 9
```

```{r three-arm-plot, fig.cap = "Three-arm MMRM with pairwise brackets."}
lplot(df3, measure ~ visit | group,
  baseline_value = 0,
  cluster_var = "subject_id",
  statistical_annotations = TRUE,
  test_method = "mmrm",
  title = "Three-Arm MMRM: Dose Response")
```

# Combined with sample size table

The MMRM annotations pair naturally with the below-axis sample size
table for a publication-ready figure.

```{r combined, fig.cap = "MMRM with sample size table and NEJM theme."}
lplot(df, measure ~ visit | group,
  baseline_value = 0,
  cluster_var = "subject_id",
  statistical_annotations = TRUE,
  test_method = "mmrm",
  show_sample_sizes = TRUE,
  sample_size_opts = list(position = "table"),
  theme = "nejm",
  title = "Efficacy Endpoint: MMRM Analysis")
```

# Comparison with pointwise tests

The table below summarizes the three `test_method` options and their
appropriate use cases.

| Method | Model | Two groups | Three+ groups | Use case |
|:-------|:------|:-----------|:-------------|:---------|
| `"parametric"` | Independent tests per visit | Welch t-test | ANOVA (omnibus) + pairwise t-tests | Exploratory, descriptive |
| `"nonparametric"` | Independent tests per visit | Wilcoxon rank-sum | Kruskal-Wallis (omnibus) + pairwise Wilcoxon | Non-normal data, ordinal endpoints |
| `"mmrm"` | Joint model across visits | Pairwise emmeans | Joint test (omnibus) + pairwise emmeans | Primary efficacy analysis, regulatory submissions |

Key advantages of MMRM over pointwise tests:

- Borrows information across timepoints through the covariance
  model, yielding more precise estimates.
- Handles missing data under the missing-at-random (MAR) assumption
  without imputation.
- Produces valid inference when the dropout pattern depends on
  observed (but not unobserved) outcomes.
- Aligns with ICH E9(R1) guidance on estimands in clinical trials.

# Requirements

The MMRM option requires two additional packages (listed in
Suggests, not Imports):

- `mmrm`: model fitting (Satterthwaite degrees of freedom by
  default)
- `emmeans`: estimated marginal means and pairwise contrasts

Install with:

```r
pak::pak(c("mmrm", "emmeans"))
```

---
*Rendered on 2026-03-17 at 11:58 PDT.*
*Source: ~/prj/sfw/01-zzlongplot/zzlongplot/vignettes/mmrm-analysis.Rmd*
