---
title: "Quickstart Guide: zzlongplot"
author: "Ronald (Ryy) G. Thomas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quickstart Guide: zzlongplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Installation

```{r install, eval=FALSE}
devtools::install_github("rgt47/zzlongplot")
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  eval = TRUE
)
devtools::load_all()
library(dplyr)
library(ggplot2)
```

## Overview

**zzlongplot** provides a flexible framework for longitudinal data
visualization in R. The package uses a formula-based interface for
creating publication-quality plots of repeated measurements over
time, with specialized support for clinical trials.


## Simulated Data

All examples use the following two-arm and three-arm datasets.

```{r data}
set.seed(42)
n_subj <- 60

subj_info <- data.frame(
  subject_id = 1:n_subj,
  treatment = rep(c("Active", "Placebo"), each = 30)
)

two_arm <- expand.grid(
  subject_id = 1:n_subj,
  visit = c(0, 4, 8, 12)
) |>
  left_join(subj_info, by = "subject_id") |>
  mutate(
    response = 50 +
      visit * ifelse(treatment == "Active", -2.5, -0.5) +
      rnorm(n(), 0, 6)
  )

subj_info3 <- data.frame(
  subject_id = 1:90,
  treatment = rep(
    c("High Dose", "Low Dose", "Placebo"), each = 30
  )
)

three_arm <- expand.grid(
  subject_id = 1:90,
  visit = c(0, 4, 8, 12)
) |>
  left_join(subj_info3, by = "subject_id") |>
  mutate(
    response = 50 +
      visit * ifelse(
        treatment == "High Dose", -3,
        ifelse(treatment == "Low Dose", -1.5, -0.5)
      ) + rnorm(n(), 0, 6)
  )
```

## Basic Usage

The main function `lplot()` uses a formula interface:
`outcome ~ time | group`.

### Observed Values with Error Bars

```{r obs-bars, fig.cap="Figure 1: Observed values with error bars (default)."}
# Figure 1
lplot(two_arm,
  response ~ visit | treatment,
  baseline_value = 0,
  caption = "Error bars represent mean +/- 1 SE.",
  title = "Observed Values: Error Bars")
```

### Observed Values with Confidence Ribbons

```{r obs-band, fig.cap="Figure 2: Observed values with confidence ribbons."}
# Figure 2
lplot(two_arm,
  response ~ visit | treatment,
  baseline_value = 0,
  error_type = "band",
  caption = "Shaded ribbons represent mean +/- 1 SE.",
  title = "Observed Values: Confidence Ribbons")
```

### Change from Baseline

```{r change, fig.cap="Figure 3: Change from baseline plot."}
# Figure 3
lplot(two_arm,
  response ~ visit | treatment,
  baseline_value = 0,
  plot_type = "change",
  caption = "Error bars represent mean change +/- 1 SE.",
  title = "Change from Baseline")
```

### Combined Plot

```{r both, fig.width=12, fig.cap="Figure 4: Side-by-side observed and change plots."}
# Figure 4
lplot(two_arm,
  response ~ visit | treatment,
  baseline_value = 0,
  plot_type = "both",
  caption = "Error bars represent mean +/- 1 SE.",
  caption2 = "Error bars represent mean change +/- 1 SE.")
```

## Sample Size Annotations

Display per-group sample sizes at each timepoint, either as
in-plot labels or as a color-coded table below the x-axis.

```{r sample-sizes, fig.cap="Figure 5: Sample sizes as point labels."}
# Figure 5
lplot(two_arm,
  response ~ visit | treatment,
  baseline_value = 0,
  show_sample_sizes = TRUE,
  caption = "Error bars represent mean +/- 1 SE. N shown per group.",
  title = "With Sample Size Labels")
```

```{r sample-table, fig.height=6, fig.cap="Figure 6: Sample size table below axis."}
# Figure 6
lplot(two_arm,
  response ~ visit | treatment,
  baseline_value = 0,
  show_sample_sizes = TRUE,
  sample_size_opts = list(position = "table"),
  caption = "Error bars represent mean +/- 1 SE.",
  title = "With Sample Size Table")
```

## Statistical Annotations

### Two-Group Comparison

Stars appear above timepoints where the adjusted p-value
crosses standard significance thresholds.

```{r stats-2g, fig.cap="Figure 7: Two-group significance stars."}
# Figure 7
lplot(two_arm,
  response ~ visit | treatment,
  baseline_value = 0,
  statistical_annotations = TRUE,
  caption = "Error bars represent mean +/- 1 SE. *p<0.05, **p<0.01, ***p<0.001 (BH-adjusted).",
  title = "Two-Group: Significance Stars")
```

### Three-Group Comparison with Pairwise Brackets

For three or more groups, pairwise brackets with significance
stars are displayed at each timepoint.

```{r stats-3g, fig.height=6, fig.cap="Figure 8: Three-group pairwise brackets."}
# Figure 8
lplot(three_arm,
  response ~ visit | treatment,
  cluster_var = "subject_id",
  baseline_value = 0,
  statistical_annotations = TRUE,
  caption = "Error bars represent mean +/- 1 SE. *p<0.05, **p<0.01, ***p<0.001 (BH-adjusted).",
  title = "Three-Group: Pairwise Brackets")
```

## Contrast Display

When using `test_method = "mmrm"`, emmeans contrast results
(LS mean difference, 95% CI, p-value) can be displayed as a
footnote or a table below the plot.

### Footnote Mode

```{r contrast-fn, eval=FALSE, fig.height=6, fig.cap="Figure 9: MMRM with contrast footnote."}
# Figure 9
lplot(two_arm,
  response ~ visit | treatment,
  baseline_value = 0,
  statistical_annotations = TRUE,
  test_method = "mmrm",
  contrast_display = "footnote",
  caption = "Error bars represent mean +/- 1 SE. MMRM contrasts shown below.",
  title = "MMRM: Contrast Footnote")
```

### Table Mode

```{r contrast-tbl, eval=FALSE, fig.height=7, fig.cap="Figure 10: MMRM with contrast table."}
# Figure 10
lplot(two_arm,
  response ~ visit | treatment,
  baseline_value = 0,
  statistical_annotations = TRUE,
  test_method = "mmrm",
  contrast_display = "table",
  caption = "Error bars represent mean +/- 1 SE. MMRM contrast table below.",
  title = "MMRM: Contrast Table")
```

## Publication Themes

### Black-and-White Print Theme

Differentiates groups by linetype and shape rather than color,
suitable for journals that print in grayscale.

```{r bw-theme, fig.cap="Figure 11: Black-and-white print theme (default)."}
# Figure 11
lplot(two_arm,
  response ~ visit | treatment,
  baseline_value = 0,
  caption = "Error bars represent mean +/- 1 SE.",
  title = "Black-and-White Theme (Default)")
```

### NEJM Theme

```{r nejm-theme, fig.cap="Figure 12: NEJM journal theme."}
# Figure 12
lplot(two_arm,
  response ~ visit | treatment,
  baseline_value = 0,
  caption = "Error bars represent mean +/- 1 SE.",
  title = "NEJM Theme") +
  theme_nejm()
```

## Tips

### Ordering Categorical Visit Labels

When the x-axis variable is character or factor (e.g., visit
names rather than numeric week numbers), R defaults to
alphabetical order. Convert to a factor with explicit levels
before calling `lplot()`:

```{r visit-ordering, eval=FALSE}
df$visit <- factor(df$visit,
  levels = c("Screening", "Baseline", "Week 4",
             "Week 8", "Week 12"))

lplot(df, response ~ visit | treatment,
  cluster_var = "subject_id",
  baseline_value = "Baseline")
```

Numeric visit variables (e.g., `visit = c(0, 4, 8, 12)`) are
sorted numerically by default and do not require this step.

## Quick Reference

### Formula Syntax

```
outcome ~ time | group
```

- `outcome`: Response variable (y-axis)
- `time`: Time variable (x-axis)
- `group`: Grouping variable (colors/lines)

### Key Parameters

| Parameter | Description | Default |
|:----------|:------------|:--------|
| `plot_type` | "obs", "change", or "both" | "obs" |
| `error_type` | "bar" or "band" | "bar" |
| `baseline_value` | Value identifying baseline | NULL |
| `show_sample_sizes` | Show N at each timepoint | FALSE |
| `statistical_annotations` | Significance annotations | FALSE |
| `test_method` | "parametric", "nonparametric", or "mmrm" | "parametric" |
| `contrast_display` | "footnote" or "table" (MMRM) | NULL |
| `theme` | "bw" for black-and-white print | "bw" |
| `cluster_var` | Subject ID column | "subject_id" |
| `facet_form` | Faceting formula (e.g., `~ site`) | NULL |

## Next Steps

- `vignette("zzlongplot_introduction")` -- Detailed introduction
- `vignette("mmrm-analysis")` -- MMRM analysis workflow
- `vignette("clinical-trials")` -- Clinical trial applications
- `vignette("cdisc-compliance")` -- CDISC-compliant workflows
- `vignette("publication-themes")` -- Publication-ready themes

---
*Rendered on 2026-04-13 at 16:32 PDT.*
*Source: ~/prj/sfw/01-zzlongplot/zzlongplot/vignettes/quickstart.Rmd*
