---
title: "Sample Size Annotations"
author: "Ronald (Ryy) G. Thomas"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: united
    highlight: tango
vignette: >
  %\VignetteIndexEntry{Sample Size Annotations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 10,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)

library(zzlongplot)
library(dplyr)
library(ggplot2)
```

# Overview

Displaying the sample size contributing to each plotted summary
is standard practice in clinical trial figures and longitudinal
research. The `show_sample_sizes` parameter in `lplot()` places
the count next to each point. The companion `sample_size_opts`
list controls appearance (font size, color, transparency) and
placement (horizontal and vertical offsets).

# Sample Data

We use two datasets throughout this vignette: one with
continuous time and two treatment groups, and one with
categorical visits and three treatment arms.

```{r data}
set.seed(42)

continuous_df <- data.frame(
  subject_id = rep(1:60, each = 4),
  week = rep(c(0, 4, 8, 12), times = 60),
  score = NA,
  arm = rep(
    c("Drug", "Placebo"), each = 4, length.out = 240
  )
)

for (s in unique(continuous_df$subject_id)) {
  rows <- continuous_df$subject_id == s
  bl <- 50 + rnorm(1, 0, 5)
  is_drug <- continuous_df$arm[rows][1] == "Drug"
  fx <- if (is_drug) c(0, 3, 6, 10) else c(0, 1, 1.5, 2)
  continuous_df$score[rows] <- bl + fx + rnorm(4, 0, 3)
}

categorical_df <- data.frame(
  subject_id = rep(1:45, each = 3),
  visit = rep(
    c("Baseline", "Month 3", "Month 6"), times = 45
  ),
  outcome = NA,
  treatment = rep(
    c("Active A", "Active B", "Placebo"),
    each = 3, length.out = 135
  )
)

for (s in unique(categorical_df$subject_id)) {
  rows <- categorical_df$subject_id == s
  bl <- 30 + rnorm(1, 0, 4)
  trt <- categorical_df$treatment[rows][1]
  fx <- switch(trt,
    "Active A" = c(0, 5, 9),
    "Active B" = c(0, 3, 5),
    "Placebo"  = c(0, 1, 2)
  )
  categorical_df$outcome[rows] <- bl + fx + rnorm(3, 0, 2)
}
```

# Default Sample Sizes

Setting `show_sample_sizes = TRUE` with no additional options
places the count to the right of each point using default
styling (size 2.8, grey40, full opacity).

```{r default}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      show_sample_sizes = TRUE,
      title = "Default Sample Size Labels",
      xlab = "Week",
      ylab = "Score")
```

# Adjusting Font Size

Larger labels may be appropriate for presentations; smaller
labels work better for dense multi-panel figures.

```{r size-large}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      show_sample_sizes = TRUE,
      sample_size_opts = list(size = 4.5),
      title = "Larger Labels (size = 4.5)",
      xlab = "Week",
      ylab = "Score")
```

```{r size-small}
lplot(categorical_df,
      outcome ~ visit | treatment,
      cluster_var = "subject_id",
      baseline_value = "Baseline",
      show_sample_sizes = TRUE,
      sample_size_opts = list(size = 2),
      title = "Smaller Labels (size = 2)",
      xlab = "Visit",
      ylab = "Outcome")
```

# Changing Color

The label color can be set to any valid R color string. Using
black increases contrast; using a muted tone keeps labels
subordinate to the data.

```{r color}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      show_sample_sizes = TRUE,
      sample_size_opts = list(color = "black"),
      title = "Black Labels",
      xlab = "Week",
      ylab = "Score")
```

```{r color-blue}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      show_sample_sizes = TRUE,
      sample_size_opts = list(color = "steelblue", size = 3.2),
      title = "Steelblue Labels",
      xlab = "Week",
      ylab = "Score")
```

# Transparency

Setting `alpha` below 1 fades the labels so they do not
compete with the plotted data.

```{r alpha}
lplot(categorical_df,
      outcome ~ visit | treatment,
      cluster_var = "subject_id",
      baseline_value = "Baseline",
      show_sample_sizes = TRUE,
      sample_size_opts = list(alpha = 0.4),
      title = "Semi-Transparent Labels (alpha = 0.4)",
      xlab = "Visit",
      ylab = "Outcome")
```

# Horizontal Placement

The `nudge_x` option controls how far the label sits from the
point along the x-axis. Positive values shift right; negative
values shift left. For continuous x variables the unit is in
data units; for categorical x variables it is a fraction of
category spacing.

```{r nudge-x}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      show_sample_sizes = TRUE,
      sample_size_opts = list(nudge_x = 0.8),
      title = "Wider Horizontal Offset (nudge_x = 0.8)",
      xlab = "Week",
      ylab = "Score")
```

```{r nudge-x-tight}
lplot(categorical_df,
      outcome ~ visit | treatment,
      cluster_var = "subject_id",
      baseline_value = "Baseline",
      show_sample_sizes = TRUE,
      sample_size_opts = list(nudge_x = 0.05),
      title = "Tight Horizontal Offset (nudge_x = 0.05)",
      xlab = "Visit",
      ylab = "Outcome")
```

# Vertical Placement

The `nudge_y` option shifts labels up (positive) or down
(negative) in data units. This is useful when error bars or
ribbons would otherwise overlap with the labels.

```{r nudge-y-up}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      show_sample_sizes = TRUE,
      sample_size_opts = list(nudge_y = 2),
      title = "Labels Shifted Up (nudge_y = 2)",
      xlab = "Week",
      ylab = "Score")
```

```{r nudge-y-down}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      show_sample_sizes = TRUE,
      sample_size_opts = list(nudge_y = -2),
      title = "Labels Shifted Down (nudge_y = -2)",
      xlab = "Week",
      ylab = "Score")
```

# Combining Options

Multiple options can be set together. The following example
uses a larger font, black color, reduced transparency, and
a downward vertical offset.

```{r combined}
lplot(categorical_df,
      outcome ~ visit | treatment,
      cluster_var = "subject_id",
      baseline_value = "Baseline",
      show_sample_sizes = TRUE,
      sample_size_opts = list(
        size = 3.5,
        color = "black",
        alpha = 0.6,
        nudge_y = -1.5
      ),
      title = "Combined: size, color, alpha, nudge_y",
      xlab = "Visit",
      ylab = "Outcome")
```

# With Change-from-Baseline Plots

Sample size labels appear on both observed and change panels
when `plot_type = "both"`. The same `sample_size_opts` apply
to both panels.

```{r change-both}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      plot_type = "both",
      show_sample_sizes = TRUE,
      sample_size_opts = list(
        size = 3, color = "grey30", alpha = 0.7
      ),
      title = "Observed Values",
      title2 = "Change from Baseline",
      xlab = "Week",
      ylab = "Score",
      ylab2 = "Score Change")
```

# With BW Print Theme

The black-and-white theme pairs well with sample size
annotations for figures destined for monochrome printing.

```{r bw-theme}
lplot(categorical_df,
      outcome ~ visit | treatment,
      cluster_var = "subject_id",
      baseline_value = "Baseline",
      theme = "bw",
      plot_type = "both",
      show_sample_sizes = TRUE,
      sample_size_opts = list(
        size = 3, color = "black", alpha = 0.5
      ),
      title = "Observed (BW)",
      title2 = "Change (BW)",
      xlab = "Visit",
      ylab = "Outcome",
      ylab2 = "Outcome Change")
```

# With Error Bands

When using ribbon-style error bands, shifting labels
vertically can prevent overlap with the shaded region.

```{r ribbons}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      error_type = "band",
      show_sample_sizes = TRUE,
      sample_size_opts = list(
        nudge_y = 3, size = 3, color = "grey20"
      ),
      title = "Error Bands with Elevated Labels",
      xlab = "Week",
      ylab = "Score")
```

# Parameter Reference

The `sample_size_opts` list accepts the following elements.
All are optional; omitted elements use their defaults.

| Option    | Default  | Description                         |
|:----------|:---------|:------------------------------------|
| `size`    | 2.8      | Font size in mm                     |
| `color`   | "grey40" | Label color (any R color)           |
| `alpha`   | 1        | Transparency, 0 (invisible) to 1    |
| `nudge_x` | auto     | Horizontal offset from point        |
| `nudge_y` | 0        | Vertical offset from point          |

When `nudge_x` is not specified, it is automatically
calculated as 3% of the x-axis range for continuous variables
or 0.15 category units for categorical variables.

# Session Info

```{r session-info}
sessionInfo()
```
