---
title: "Sample Size Annotations"
author: "Ronald (Ryy) G. Thomas"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Sample Size Annotations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  dpi = 96,
  fig.retina = 1,
  echo = TRUE,
  fig.width = 10,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)

library(zzlongplot)
library(dplyr)
library(ggplot2)
```

# Overview

Displaying the sample size contributing to each plotted summary
is standard practice in clinical trial figures and longitudinal
research. The `show_sample_sizes` parameter in `lplot()` places
the count next to each point. The companion `sample_size_opts`
list controls appearance (font size, color, transparency) and
placement (horizontal and vertical offsets).

# Sample Data

We use two datasets throughout this vignette: one with
continuous time and two treatment groups, and one with
categorical visits and three treatment arms.

```{r data}
set.seed(42)

continuous_df <- data.frame(
  subject_id = rep(1:60, each = 4),
  week = rep(c(0, 4, 8, 12), times = 60),
  score = NA,
  arm = rep(
    c("Drug", "Placebo"), each = 4, length.out = 240
  )
)

for (s in unique(continuous_df$subject_id)) {
  rows <- continuous_df$subject_id == s
  bl <- 50 + rnorm(1, 0, 5)
  is_drug <- continuous_df$arm[rows][1] == "Drug"
  fx <- if (is_drug) c(0, 3, 6, 10) else c(0, 1, 1.5, 2)
  continuous_df$score[rows] <- bl + fx + rnorm(4, 0, 3)
}

categorical_df <- data.frame(
  subject_id = rep(1:45, each = 3),
  visit = rep(
    c("Baseline", "Month 3", "Month 6"), times = 45
  ),
  outcome = NA,
  treatment = rep(
    c("Active A", "Active B", "Placebo"),
    each = 3, length.out = 135
  )
)

for (s in unique(categorical_df$subject_id)) {
  rows <- categorical_df$subject_id == s
  bl <- 30 + rnorm(1, 0, 4)
  trt <- categorical_df$treatment[rows][1]
  fx <- switch(trt,
    "Active A" = c(0, 5, 9),
    "Active B" = c(0, 3, 5),
    "Placebo"  = c(0, 1, 2)
  )
  categorical_df$outcome[rows] <- bl + fx + rnorm(3, 0, 2)
}
```

# Default Sample Sizes

Setting `show_sample_sizes = TRUE` with no additional options
places the count to the right of each point using default
styling (size 2.8, grey40, full opacity).

```{r default}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      show_sample_sizes = TRUE,
      title = "Default Sample Size Labels",
      xlab = "Week",
      ylab = "Score")
```

# Adjusting Font Size

Larger labels may be appropriate for presentations; smaller
labels work better for dense multi-panel figures.

```{r size-large}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      show_sample_sizes = TRUE,
      sample_size_opts = list(size = 4.5),
      title = "Larger Labels (size = 4.5)",
      xlab = "Week",
      ylab = "Score")
```

```{r size-small}
lplot(categorical_df,
      outcome ~ visit | treatment,
      cluster_var = "subject_id",
      baseline_value = "Baseline",
      show_sample_sizes = TRUE,
      sample_size_opts = list(size = 2),
      title = "Smaller Labels (size = 2)",
      xlab = "Visit",
      ylab = "Outcome")
```

# Changing Color

The label color can be set to any valid R color string. Using
black increases contrast; using a muted tone keeps labels
subordinate to the data.

```{r color}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      show_sample_sizes = TRUE,
      sample_size_opts = list(color = "black"),
      title = "Black Labels",
      xlab = "Week",
      ylab = "Score")
```

```{r color-blue}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      show_sample_sizes = TRUE,
      sample_size_opts = list(color = "steelblue", size = 3.2),
      title = "Steelblue Labels",
      xlab = "Week",
      ylab = "Score")
```

# Transparency

Setting `alpha` below 1 fades the labels so they do not
compete with the plotted data.

```{r alpha}
lplot(categorical_df,
      outcome ~ visit | treatment,
      cluster_var = "subject_id",
      baseline_value = "Baseline",
      show_sample_sizes = TRUE,
      sample_size_opts = list(alpha = 0.4),
      title = "Semi-Transparent Labels (alpha = 0.4)",
      xlab = "Visit",
      ylab = "Outcome")
```

# Horizontal Placement

The `nudge_x` option controls how far the label sits from the
point along the x-axis. Positive values shift right; negative
values shift left. For continuous x variables the unit is in
data units; for categorical x variables it is a fraction of
category spacing.

```{r nudge-x}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      show_sample_sizes = TRUE,
      sample_size_opts = list(nudge_x = 0.8),
      title = "Wider Horizontal Offset (nudge_x = 0.8)",
      xlab = "Week",
      ylab = "Score")
```

```{r nudge-x-tight}
lplot(categorical_df,
      outcome ~ visit | treatment,
      cluster_var = "subject_id",
      baseline_value = "Baseline",
      show_sample_sizes = TRUE,
      sample_size_opts = list(nudge_x = 0.05),
      title = "Tight Horizontal Offset (nudge_x = 0.05)",
      xlab = "Visit",
      ylab = "Outcome")
```

# Vertical Placement

The `nudge_y` option shifts labels up (positive) or down
(negative) in data units. This is useful when error bars or
ribbons would otherwise overlap with the labels.

```{r nudge-y-up}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      show_sample_sizes = TRUE,
      sample_size_opts = list(nudge_y = 2),
      title = "Labels Shifted Up (nudge_y = 2)",
      xlab = "Week",
      ylab = "Score")
```

```{r nudge-y-down}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      show_sample_sizes = TRUE,
      sample_size_opts = list(nudge_y = -2),
      title = "Labels Shifted Down (nudge_y = -2)",
      xlab = "Week",
      ylab = "Score")
```

# Combining Options

Multiple options can be set together. The following example
uses a larger font, black color, reduced transparency, and
a downward vertical offset.

```{r combined}
lplot(categorical_df,
      outcome ~ visit | treatment,
      cluster_var = "subject_id",
      baseline_value = "Baseline",
      show_sample_sizes = TRUE,
      sample_size_opts = list(
        size = 3.5,
        color = "black",
        alpha = 0.6,
        nudge_y = -1.5
      ),
      title = "Combined: size, color, alpha, nudge_y",
      xlab = "Visit",
      ylab = "Outcome")
```

# With Change-from-Baseline Plots

Sample size labels appear on both observed and change panels
when `plot_type = "both"`. The same `sample_size_opts` apply
to both panels.

```{r change-both}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      plot_type = "both",
      show_sample_sizes = TRUE,
      sample_size_opts = list(
        size = 3, color = "grey30", alpha = 0.7
      ),
      title = "Observed Values",
      title2 = "Change from Baseline",
      xlab = "Week",
      ylab = "Score",
      ylab2 = "Score Change")
```

# With BW Print Theme

The black-and-white theme pairs well with sample size
annotations for figures destined for monochrome printing.

```{r bw-theme}
lplot(categorical_df,
      outcome ~ visit | treatment,
      cluster_var = "subject_id",
      baseline_value = "Baseline",
      theme = "bw",
      plot_type = "both",
      show_sample_sizes = TRUE,
      sample_size_opts = list(
        size = 3, color = "black", alpha = 0.5
      ),
      title = "Observed (BW)",
      title2 = "Change (BW)",
      xlab = "Visit",
      ylab = "Outcome",
      ylab2 = "Outcome Change")
```

# With Error Bands

When using ribbon-style error bands, shifting labels
vertically can prevent overlap with the shaded region.

```{r ribbons}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      error_type = "band",
      show_sample_sizes = TRUE,
      sample_size_opts = list(
        nudge_y = 3, size = 3, color = "grey20"
      ),
      title = "Error Bands with Elevated Labels",
      xlab = "Week",
      ylab = "Score")
```

# A Table Below the Axis

Everything above places a label beside its point, which is
`position = "point"`, the default. With many timepoints those
labels crowd the markers. `position = "table"` instead lays
the counts out in a row per group beneath the panel, in the
group colors, the way a CONSORT-style figure usually reports
them.

```{r ss-table}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      show_sample_sizes = TRUE,
      sample_size_opts = list(position = "table"),
      title = "Counts as a table below the axis",
      xlab = "Week",
      ylab = "Score")
```

By default the table names each row at its left edge and
hides the legend, on the assumption that those row names
identify the groups. That is not always what you want. If
the figure needs its legend, or the row names crowd the
first timepoint, drop the names and keep the legend
instead.

```{r ss-table-legend}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      show_sample_sizes = TRUE,
      sample_size_opts = list(position = "table",
                              show_group_labels = FALSE,
                              legend = "keep"),
      title = "Row names dropped, legend kept",
      xlab = "Week",
      ylab = "Score")
```

## Where the rows are drawn

The table needs vertical space, and there are two places to
take it from. `region = "margin"`, the default, holds the
panel to the data range and reserves plot margin underneath
it. That is the cleaner result when the figure has room.

`region = "panel"` instead expands the y scale so the rows
sit inside the panel. Prefer it when the figure is already
tall on furniture -- a multi-line subtitle, a bottom legend,
a short device -- where a reserved margin can squeeze the
panel to a sliver.

```{r ss-region}
lplot(continuous_df,
      score ~ week | arm,
      cluster_var = "subject_id",
      baseline_value = 0,
      show_sample_sizes = TRUE,
      sample_size_opts = list(position = "table",
                              show_group_labels = FALSE,
                              legend = "keep",
                              region = "panel"),
      title = "Counts inside the panel",
      subtitle = paste("region = 'panel' spends vertical space",
                       "inside the panel rather than reserving",
                       "margin beneath it."),
      xlab = "Week",
      ylab = "Score")
```

## Faceted figures

Under faceting each panel reports its own counts. This
matters whenever the panels differ in size, which is the
usual reason for faceting in the first place.

```{r ss-facets}
continuous_df$stratum <- ifelse(continuous_df$subject_id %%
                                  4 == 0, "Stratum B", "Stratum A")

lplot(continuous_df,
      score ~ week | arm,
      facet_form = ~ stratum,
      cluster_var = "subject_id",
      baseline_value = 0,
      show_sample_sizes = TRUE,
      sample_size_opts = list(position = "table",
                              show_group_labels = FALSE,
                              legend = "keep",
                              region = "panel"),
      title = "Each panel reports its own n",
      xlab = "Week",
      ylab = "Score")
```

Before version 0.3.0 the table resolved its counts with a
first-match lookup and carried no facet information, so the
first panel's sample sizes were repeated in every panel. A
faceted figure with unequal panels therefore stated sample
sizes that were not its own. Regenerate any such figure
produced with an earlier version.

# Parameter Reference

The `sample_size_opts` list accepts the following elements.
All are optional; omitted elements use their defaults.

| Option              | Default  | Applies to | Description                                      |
|:--------------------|:---------|:-----------|:-------------------------------------------------|
| `position`          | "point"  | both       | "point" beside each marker, or "table" below the axis |
| `size`              | 2.8      | both       | Font size in mm                                  |
| `alpha`             | 1        | both       | Transparency, 0 (invisible) to 1                 |
| `color`             | "grey40" | point      | Label color (any R color); table mode uses group colors |
| `nudge_x`           | auto     | point      | Horizontal offset from point                     |
| `nudge_y`           | 0        | point      | Vertical offset from point                       |
| `gap`               | 0.18     | table      | Fraction of the y-range between panel and first row |
| `row_height`        | 0.06     | table      | Fraction of the y-range between rows             |
| `label_size`        | `size`   | table      | Font size for the row names                      |
| `label_offset`      | 0.08     | table      | Horizontal offset of the row names               |
| `show_group_labels` | TRUE     | table      | Print the group name at the left of each row     |
| `legend`            | "none"   | table      | "none" hides the legend, "keep" retains it       |
| `region`            | "margin" | table      | "margin" reserves space below the panel, "panel" expands the y scale |

When `nudge_x` is not specified, it is automatically
calculated as 3% of the x-axis range for continuous variables
or 0.15 category units for categorical variables.

`show_group_labels` and `legend` are usually set together:
turning the row names off leaves nothing to identify the
rows unless the legend is kept.

# Session Info

```{r session-info}
sessionInfo()
```
